---
title: "Hands on Exercise 2"
author: "Chandru"
---

Getting Started

In this hands-on exercise, the key R package use is tmap package in R. Beside tmap package, five other R packages will be used. They are:

-   [**readr**](https://readr.tidyverse.org/) for importing delimited text file,

-   [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,

-   [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data and

-   [**sf**](https://cran.r-project.org/web/packages/sf/index.html) for handling geospatial data.

-   [**rvest**](https://rvest.tidyverse.org/) for scraping (or harvesting) data from web pages.

Among the five packages, readr, tidyr and dplyr are part of tidyverse package.

The code chunk below will be used to install and load these packages in RStudio.



```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

Importing Geospatial Data into R

The code chunk below uses the st_read() function of sf package to import MP14_SUBZONE_WEB_PL shapefile into R as a simple feature data frame called mpsz.

```{r}
mpsz <- st_read("Data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```
Function to extract values from the HTML description

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

You can examine the content of mpsz by using the code chunk below.

```{r}
mpsz
```

Next, we will import respopagesextod2024.csv file into RStudio and save the file into an tibble dataframe called popdata.

The task will be performed by using read_csv() function of readr package as shown in the code chunk below.

```{r}
popdata <- read_csv("Data/aspatial/respopagesextod2024.csv")
```

Before a thematic map can be prepared, you are required to prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

YOUNG: age group 0 to 4 until age groyup 20 to 24,
ECONOMY ACTIVE: age group 25-29 until age group 60-64,
AGED: age group 65 and above,
TOTAL: all age group, and
DEPENDENCY: the ratio between young and aged against economy active group


Data Wrangling

The following data wrangling and transformation functions will be used:

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```
Joining the attribute data and geospatial data

Before we can perform the georelational join, one extra step is required to convert the values in PA and SZ fields to uppercase. This is because the values of PA and SZ fields are made up of upper- and lowercase. On the other, hand the SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, left_join() of dplyr is used to join the geographical data and attribute table using planning subzone name e.g. SUBZONE_N and SZ as the common identifier.

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

Thing to learn from the code chunk above:

left_join() of dplyr package is used with mpsz simple feature data frame as the left data table is to ensure that the output will be a simple features data frame.

```{r}
write_rds(mpsz_pop2024, "Data/rds/mpsz_pop2024.rds")
```

Choropleth Mapping Geospatial Data Using tmap

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

Two approaches can be used to prepare thematic map using tmap, they are:

Plotting a thematic map quickly by using qtm().
Plotting highly customisable thematic map by using tmap elements.


Plotting a choropleth map quickly by using qtm()

The easiest and quickest to draw a choropleth map using tmap is using qtm(). It is concise and provides a good default visualisation in many cases.

The code chunk below will draw a cartographic standard choropleth map as shown below.

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2024, 
    fill = "DEPENDENCY")
```

Creating a choropleth map by using tmap’s elements

Despite its usefulness of drawing a choropleth map quickly and easily, the disadvantge of qtm() is that it makes aesthetics of individual layers harder to control. To draw a high quality cartographic choropleth map as shown in the figure below, tmap’s drawing elements should be used.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

