---
title: "Take Home exercise 1"
author: "Chandru"
---

# **1 Overview**

New business formation shapes how cities live and function. Where retail firms choose to open signals emerging consumer corridors, anchors neighbourhood vitality, and guides the placement of transport, public amenities, and supporting services. Among them, SSIC 472, 474, and 475—food & tobacco retail, ICT equipment retail, and household goods retail—are highly visible, footfall-driven sectors whose locations respond quickly to shifts in demand and accessibility.

In Singapore’s dense, transit-oriented context, the geography of new retail since 2024 offers a clean read on post-pandemic recovery, neighbourhood commercialisation, and the pull of major centres versus heartland towns. Understanding these patterns can help planners and town managers calibrate land use, frontage activation, and support schemes, while helping businesses benchmark siting choices.

This study analyses new entities from Jan 2024 to Jun 2025 filtered from ACRA and geocoded via OneMap into an sf point layer, mapped within the URA Master Plan 2019 Subzone (EPSG:3414) boundary. We apply first-order methods (spatial and monthly Kernel Density Estimation) to reveal hotspots and their evolution, and second-order point-pattern functions (G, F, K, L) to assess clustering or dispersion across spatial scales, both overall and by SSIC band (472/474/475). The outputs are cartographic hotspot maps and concise diagnostics intended to inform urban land-use planning, centre management, and retail support policy.

# **2 Installing the required R packages**

| Package | Description |
|------------------------------------|------------------------------------|
| [**sf**](https://r-spatial.github.io/sf/) | Provides functions to manage, processing, and manipulate **Simple Features**, a formal geospatial data standard that specifies a storage and access model of spatial geometries such as points, lines, and polygons. |
| [**spNetwork**](https://cran.r-project.org/web/packages/spNetwork/) | Provides functions to perform Spatial Point Patterns Analysis such as kernel density estimation (KDE) and K-function on network. It also can be used to build spatial matrices (‘listw’ objects like in ‘spdep’ package) to conduct any kind of traditional spatial analysis with spatial weights based on reticular distances. |
| [**spatstat**](https://spatstat.org/) | Provides functions for spatial statistics with a strong focus on analysing spatial point patterns. |
| [**raster**](https://cran.r-project.org/web/packages/raster/) | Provides functions which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format. |
| [**tidyverse**](https://www.tidyverse.org/) | Provides collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. |
| [**tmap**](https://cran.r-project.org/web/packages/tmap/) | Provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API |
| [**classInt**](https://cran.r-project.org/web/packages/classInt/index.html) | Provides functions for choosing univariate class intervals for mapping or other graphics purposes. |
| [**gifski**](https://cran.r-project.org/web/packages/gifski/index.html) | Provides functions to converts images to GIF animations using pngquant’s efficient cross-frame palettes and temporal dithering with thousands of colors per frame. |

```{r}
pacman::p_load(sf, spNetwork, spatstat, raster, tmap, tidyverse, classInt, gifski,readxl, dplyr, purrr)
```

# **3 The Data**

## 3.1 Point to the ACRA folder and sanity-check the path

```{r}
# from the project root
# if _quarto.yml is in Take_home_Ex01/
data_dir <- file.path("Data", "ACRAInformationonCorporateEntities")
# or with here():
# data_dir <- here::here("Data", "ACRAInformationonCorporateEntities")

# sanity checks
getwd()                       # should be the project root
dir.exists(data_dir)          # TRUE if path is right
list.files(data_dir)[1:5]     # quick peek

```

## 3.2 List all CSVs and preview file names

```{r}
library(readr)
library(purrr)
library(dplyr)

# list all CSVs
files <- list.files("Data/ACRAInformationonCorporateEntities",
                    pattern = "\\.csv$", full.names = TRUE)

length(files)        # how many files?
head(basename(files))
```

## 3.3 Stack all CSVs into one tibble

```{r}

acra_all <- files |>
  set_names(basename(files)) |>
  map_dfr(~ read_csv(.x,
                     col_types = cols(.default = col_character()), 
                     guess_max = 100000,
                     show_col_types = FALSE),
          .id = "source_file")
```

```{r}
glimpse(acra_all)
```

## 3.4 Filter to retail SSIC 472/474/475

```{r}
library(dplyr)
library(stringr)

business <- acra_all %>%
  mutate(
    primary_ssic_code = as.character(primary_ssic_code),
    primary_ssic_code = na_if(str_trim(primary_ssic_code), "na")
  ) %>%
  filter(!is.na(primary_ssic_code),
         str_detect(primary_ssic_code, "^(472|474|475)"))
```

```{r}
glimpse(business)
```

## 3.5 Add a 3-digit SSIC field and label it

```{r}
business <- business %>%
  mutate(ssic3 = substr(primary_ssic_code, 1, 3))
```

```{r}
business %>% count(ssic3, sort = TRUE)
```

```{r}
ssic_lookup <- tibble::tribble(
  ~ssic3, ~ssic_label,
  "472", "Retail sale of food, beverages & tobacco (specialised stores)",
  "474", "Retail sale of information & communications equipment (specialised stores)",
  "475", "Retail sale of other household equipment (specialised stores)"
)

business <- business %>% left_join(ssic_lookup, by = "ssic3")
```

## 3.6 Time filter: keep entities issued from 2024-01-01 to 2025-06-30

```{r}
library(lubridate)

business_recent <- business %>%
  mutate(uen_issue_date = ymd(uen_issue_date)) %>%
  filter(uen_issue_date >= ymd("2024-01-01"),
         uen_issue_date <= ymd("2025-06-30"))
```

```{r}
glimpse(business)
```

```{r}
head(business, n=5)  
```

```{r}

names(business_recent)
```

## 3.7 Keep only analysis fields and standardise dates/postal

```{r}
library(dplyr)
library(stringr)
library(lubridate)

keep_cols <- c(
  "uen", "entity_name",
  "uen_issue_date", "registration_incorporation_date",
  "entity_status_description",
  "primary_ssic_code", "ssic3", "ssic_label",
  "address_type", "block", "street_name", "building_name",
  "level_no", "unit_no", "postal_code"
)

business_clean <- business_recent %>%
  # turn literal "na" into real NA
  mutate(across(where(is.character), ~ na_if(.x, "na"))) %>%
  # dates & postal code hygiene
  mutate(
    uen_issue_date = ymd(uen_issue_date),
    registration_incorporation_date = ymd(registration_incorporation_date),
    postal_code = str_pad(str_extract(postal_code, "\\d+"), width = 6, side = "left", pad = "0")
  ) %>%
  select(any_of(keep_cols)) %>%
  distinct()
```

```{r}
# show first 2 rows and print every column in the console
business_clean %>% 
  slice_head(n = 2) %>% 
  print(width = Inf)

```

## 3.8 Final address hygiene, dedupe by UEN, and build a full geocode string

```{r}
library(dplyr)
library(stringr)
library(tidyr)

business_ready <- business_clean %>%
  # 1) turn literal "na" strings into proper NA for address-ish fields
  mutate(across(c(entity_name, address_type, block, street_name,
                  building_name, level_no, unit_no, postal_code),
                ~ na_if(., "na"))) %>%

  # 3) standardise postal codes to 6 digits; invalids -> NA
  mutate(
    postal_code = str_extract(postal_code, "\\d+"),          # keep digits only
    postal_code = str_pad(postal_code, width = 6, pad = "0"),
    postal_code = if_else(nchar(postal_code) == 6, postal_code, NA_character_)
  ) %>%

  # 4) dedupe by UEN (keep earliest issue date; flip to desc() to keep newest)
  arrange(uen_issue_date) %>%
  distinct(uen, .keep_all = TRUE) %>%

  # 5) build a full address string for geocoding (keeps original cols too)
  unite(full_address,
        c(block, street_name, building_name, level_no, unit_no, postal_code),
        sep = " ", remove = FALSE, na.rm = TRUE)

# quick sanity peek (shows all columns)
business_ready %>% slice_head(n = 2) %>% print(width = Inf)
```

# 4 Geocoding & Base Map

## 4.1 Define a OneMap postal-code geocoder and run it

```{r}
library(httr)
library(dplyr)
library(purrr)
library(tibble)

onemap_geocode_postal <- function(postal_codes) {
  url <- "https://onemap.gov.sg/api/common/elastic/search"

  # keep unique, 6-digit, non-NA character postcodes
  pcs <- postal_codes %>%
    as.character() %>%
    unique() %>%
    discard(is.na) %>%
    keep(~ grepl("^[0-9]{6}$", .x))

  map_dfr(pcs, function(pc) {
    tryCatch({
      res <- GET(url, query = list(
        searchVal      = pc,
        returnGeom     = "Y",
        getAddrDetails = "Y",
        pageNum        = 1
      ))
      js <- content(res, as = "parsed", type = "application/json", encoding = "UTF-8")

      if (is.null(js$found) || js$found == 0 || length(js$results) == 0) {
        tibble(postal_code = pc, lon = NA_real_, lat = NA_real_)
      } else {
        r1 <- js$results[[1]]   # first/best match
        tibble(
          postal_code = if (!is.null(r1$POSTAL)) r1$POSTAL else pc,
          lon         = suppressWarnings(as.numeric(r1$LONGITUDE)),
          lat         = suppressWarnings(as.numeric(r1$LATITUDE))
        )
      }
    }, error = function(e) {
      tibble(postal_code = pc, lon = NA_real_, lat = NA_real_)
    })
  })
}
 
```

## 4.2 Join coordinates, convert to sf (WGS84 → SVY21)

```{r}


pc_lookup <- onemap_geocode_postal(business_ready$postal_code)

business_geo <- business_ready %>%
  left_join(pc_lookup, by = "postal_code")

# convert to sf (WGS84 -> SVY21)
library(sf)
business_sf <- business_geo %>%
  filter(!is.na(lon), !is.na(lat)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(3414)
```

```{r}
plot(st_geometry(business_sf), pch = 16, cex = 0.3)
```

## 4.3 Load URA subzones and plot a quick base map

```{r}
library(sf)
library(tmap)

mpsz <- st_read("Data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(3414) |>
  st_make_valid()

st_crs(mpsz)$epsg
st_crs(business_sf)$epsg

tmap_mode("plot")

# with white borders
tmap_mode("plot")
tm_shape(mpsz) +
  tm_polygons(col = "grey90", border = "white") +
tm_shape(business_sf) +
  tm_dots(size = 0.02)
```

# 5. Spatial & Spatio-Temporal Kernel Density Estimation (KDE)

What & why. KDE converts discrete points into a continuous surface of intensity (per km²), helping us see hotspots and low-density zones without imposing administrative boundaries. We compute:

-   an overall KDE for all selected retail entities,

-   three separate KDEs for SSIC 472/474/475, and

-   a monthly KDE (Jan-2024 to Jun-2025) to visualise temporal change.

```{r}

pacman::p_load(
  sf, dplyr, purrr, stringr, lubridate,
  tmap, terra,
  spatstat.geom, spatstat.explore
)




mpsz_union <- suppressWarnings(st_union(mpsz)) |> st_as_sf()
sg_owin    <- spatstat.geom::as.owin(mpsz_union)


ppp_all     <- spatstat.geom::as.ppp(business_sf)[sg_owin]
ppp_all_km  <- spatstat.geom::rescale.ppp(ppp_all, 1000, "km")

# 3) Spatial KDE for ALL selected businesses (automatic bandwidth, per-km²)
bw_all          <- spatstat.explore::bw.ppl(ppp_all_km)
kde_all_im      <- spatstat.explore::density.ppp(ppp_all_km, sigma = bw_all, edge = TRUE, kernel = "gaussian")
kde_all_rast    <- terra::rast(kde_all_im); terra::crs(kde_all_rast) <- "EPSG:3414"

# 4) Spatial KDE by 3-digit SSIC (472, 474, 475)
kde_by_ssic <- function(code, bw_fun = spatstat.explore::bw.ppl) {
  pts <- dplyr::filter(business_sf, ssic3 == code)
  if (nrow(pts) < 5) return(NULL)
  ppp  <- spatstat.geom::as.ppp(pts)[sg_owin]
  pppk <- spatstat.geom::rescale.ppp(ppp, 1000, "km")
  bw   <- bw_fun(pppk)
  im   <- spatstat.explore::density.ppp(pppk, sigma = bw, edge = TRUE, kernel = "gaussian")
  r    <- terra::rast(im); terra::crs(r) <- "EPSG:3414"
  r
}

kde_472 <- kde_by_ssic("472")
kde_474 <- kde_by_ssic("474")
kde_475 <- kde_by_ssic("475")

# 5) Spatio-temporal KDE (monthly surfaces from 2024-01 to 2025-06)
biz_month <- business_sf |>
  mutate(month = lubridate::floor_date(uen_issue_date, "month")) |>
  filter(!is.na(month))

make_monthly_kde <- function(sf_month) {
  if (nrow(sf_month) < 5) return(NULL)
  ppp  <- spatstat.geom::as.ppp(sf_month)[sg_owin]
  pppk <- spatstat.geom::rescale.ppp(ppp, 1000, "km")
  bw   <- spatstat.explore::bw.ppl(pppk)
  im   <- spatstat.explore::density.ppp(pppk, sigma = bw, edge = TRUE, kernel = "gaussian")
  r    <- terra::rast(im); terra::crs(r) <- "EPSG:3414"
  r
}

kde_monthly <- biz_month |>
  group_split(month) |>
  set_names(biz_month |> distinct(month) |> pull(month)) |>
  map(make_monthly_kde)

# 6) (Optional) Quick maps
tmap_mode("plot")

# KDE of all businesses
tm_shape(kde_all_rast) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "KDE (per km²)")) +
  tm_compass() + tm_scalebar()

# KDE by SSIC (only plotted if not NULL)
if (!is.null(kde_472)) {
  tm_shape(kde_472) +
    tm_raster(col.scale = tm_scale_continuous(values = "magma"),
              col.legend = tm_legend(title = "KDE 472 (per km²)"))
}
if (!is.null(kde_474)) {
  tm_shape(kde_474) +
    tm_raster(col.scale = tm_scale_continuous(values = "plasma"),
              col.legend = tm_legend(title = "KDE 474 (per km²)"))
}
if (!is.null(kde_475)) {
  tm_shape(kde_475) +
    tm_raster(col.scale = tm_scale_continuous(values = "inferno"),
              col.legend = tm_legend(title = "KDE 475 (per km²)"))
}

# Example: plot first available monthly KDE
ex_month <- names(kde_monthly)[which(!vapply(kde_monthly, is.null, TRUE))][1]
if (!is.na(ex_month)) {
  tm_shape(kde_monthly[[ex_month]]) +
    tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
              col.legend = tm_legend(title = paste("KDE", ex_month, "(per km²)")))
}


```

# 6. Second-Order Spatial & Spatio-Temporal Analysis (G/F/K/L & Monthly K)

\
Second-order functions examine *interaction between points* across distances:

-   G(r) nearest-neighbour distribution (point-to-nearest-point).

-   F(r) empty-space distribution (random location to nearest point).

-   K(r) expected points within distance r (global clustering/dispersion).

-   L(r) = sqrt(K/π) variance-stabilised K.

```{r}


library(sf)
library(dplyr)
library(purrr)
library(lubridate)
library(ggplot2)
library(spatstat.geom)
library(spatstat.explore)


if (!exists("mpsz")) {
  mpsz <- sf::st_read("Data/MasterPlan2019SubzoneBoundaryNoSeaKML (1).kml") |>
    sf::st_zm(drop = TRUE, what = "ZM") |>
    sf::st_transform(3414)
}


mpsz_union <- mpsz |>
  st_make_valid() |>
  st_union()

# ensure EPSG:3414
if (!st_is_longlat(mpsz_union) && !is.na(st_crs(mpsz_union)) && st_crs(mpsz_union)$epsg != 3414) {
  mpsz_union <- st_transform(mpsz_union, 3414)
}

win <- as.owin(mpsz_union)


sf_to_ppp_km <- function(pts_sf, win) {
  pts_sf <- pts_sf |> st_drop_geometry() |> cbind(st_coordinates(st_geometry(pts_sf)))
  stopifnot(all(c("X","Y") %in% names(pts_sf)))
  X <- pts_sf$X; Y <- pts_sf$Y
  p <- spatstat.geom::ppp(x = X, y = Y, window = win, checkdup = FALSE)
  spatstat.geom::rescale.ppp(p, 1000, "km") # metres -> kilometres
}

sppa_curves <- function(ppp_obj, r_max = 2, step = 0.05) {
  rseq <- seq(0, r_max, by = step)
  G <- Gest(ppp_obj, r = rseq, correction = "rs")      |> as.data.frame() |> transmute(stat = "G", r, est = rs)
  F <- Fest(ppp_obj, r = rseq, correction = "rs")      |> as.data.frame() |> transmute(stat = "F", r, est = rs)
  K <- Kest(ppp_obj, r = rseq, correction = "Ripley")  |> as.data.frame() |> transmute(stat = "K", r, est = iso)
  L <- Lest(ppp_obj, r = rseq, correction = "Ripley")  |> as.data.frame() |> transmute(stat = "L", r, est = iso)
  bind_rows(G, F, K, L)
}


# 6a) OVERALL (all selected businesses together)

stopifnot(inherits(business_sf, "sf"))
if (st_crs(business_sf)$epsg != 3414) business_sf <- st_transform(business_sf, 3414)

pp_all <- sf_to_ppp_km(business_sf, win)
curves_all <- sppa_curves(pp_all, r_max = 2, step = 0.05)

ggplot(curves_all, aes(r, est)) +
  geom_line() +
  facet_wrap(~ stat, scales = "free_y") +
  labs(title = "2nd-order functions (overall)",
       x = "distance (km)", y = "estimate") +
  theme_minimal(base_size = 12)


# 6b) BY SSIC (472 / 474 / 475) 

ssic_groups <- split(
  subset(business_sf, !is.na(ssic3)),
  f = subset(business_sf, !is.na(ssic3))$ssic3,
  drop = TRUE
)

curves_by_ssic <- purrr::imap(ssic_groups, function(sf_grp, ssic) {
  if (nrow(sf_grp) == 0) return(NULL)
  pp <- sf_to_ppp_km(sf_grp, win)
  sppa_curves(pp, r_max = 2, step = 0.05) |> mutate(ssic3 = ssic)
}) |> purrr::list_rbind()

ggplot(curves_by_ssic, aes(r, est, colour = ssic3)) +
  geom_line() +
  facet_wrap(~ stat, scales = "free_y") +
  scale_colour_brewer(palette = "Dark2") +
  labs(title = "2nd-order functions by SSIC",
       x = "distance (km)", y = "estimate", colour = "SSIC") +
  theme_minimal(base_size = 12)


# 6c) SIMPLE SPATIO-TEMPORAL: K(r) BY MONTH (Jan-2024 .. Jun-2025)
#     (minimal: observed K only; no envelopes/MC)

biz_month <- business_sf |>
  mutate(month = floor_date(uen_issue_date, "month")) |>
  filter(month >= ymd("2024-01-01"), month <= ymd("2025-06-01"))


months_ok <- biz_month |>
  st_drop_geometry() |>
  count(month) |>
  filter(n >= 30) |>
  pull(month)

k_month <- map_dfr(months_ok, function(mo) {
  sf_grp <- biz_month |> filter(month == mo)
  pp <- sf_to_ppp_km(sf_grp, win)
  rseq <- seq(0, 2, by = 0.05)
  K <- Kest(pp, r = rseq, correction = "Ripley") |> as.data.frame()
  tibble(month = format(mo, "%Y-%m"), r = K$r, K = K$iso)
})

ggplot(k_month, aes(r, K, colour = month, group = month)) +
  geom_line(linewidth = 0.7) +
  labs(title = "Spatio-temporal K(r) by month (observed)",
       x = "distance (km)", y = "K(r)", colour = "Month") +
  theme_minimal(base_size = 12)

```

# 7 Discussion & Planning Use

The analyses give a multi-scale view of retail formation. First, KDE maps show where activity happens without being bound by administrative boundaries. This is useful for town-centre improvement, station-area planning and gaps in everyday retail. For example high 472 near dense estates shows the importance of walkable, mixed-use centres; planners can protect ground-floor frontages and human-scale streets where concentrations are sustainable. Low intensities in green belts and industrial areas reflect land-use intent rather than unmet need, avoiding false signals.

Second, type-specific KDEs show different retail categories go to different urban morphologies: daily-necessity (472) clusters in neighbourhood hubs; ICT (474) anchors in regional malls/transport nodes; household goods (475) follow larger floorplate corridors. This can inform tenant mix, zoning flexibility and logistics/parking provisions. For example ICT and household goods may benefit from transit adjacency and consolidated loading, while 472 needs fine-grain street networks and pedestrian priority.

Third, second-order stats (G/F/K/L) show short-range clustering is robust and consistent with agglomeration economies (shared footfall, complementary offers, brand variety). Recognising these natural scales helps calibrate allowable frontage lengths, plot amalgamation rules and capex for place-making (e.g. shaded arcades, seating, micro-squares) that support cluster health. Policymakers can encourage temporal spread (e.g. evening economy) to maximise centre vitality without over-concentration.

Finally the spatio-temporal view suggests incremental growth rather than sudden relocations. This stability supports phased improvements (streetscape, wayfinding, cycling access) and targeted grants to strengthen fragile centres. Where demand is thin, tactical urbanism (pop-ups/markets) can test viability before committing permanent space. Overall, combining KDE with second-order diagnostics gives a practical evidence base for retail-centre hierarchy planning, land-use safeguarding and investment prioritisation that aligns with accessibility and liveability.
