---
title: "Hands on exercise 8"
---

# **8  Spatial Weights and Applications**

## **8.1 Overview**

In this hands-on exercise, you will learn how to compute spatial weights using R. By the end to this hands-on exercise, you will be able to:

-   import geospatial data using appropriate function(s) of **sf** package,

-   import csv file using appropriate function of **readr** package,

-   perform relational join using appropriate join function of **dplyr** package,

-   compute spatial weights using appropriate functions of **spdep** package, and

-   calculate spatially lagged variables using appropriate functions of **spdep** package.

## **8.2 The Study Area and Data**

Two data sets will be used in this hands-on exercise, they are:

-   Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

-   *Hunan_2012.csv*: This csv file contains selected Hunan’s local development indicators in 2012.

### **8.2.1 Getting Started**

Before we get started, we need to ensure that **spdep**, **sf**, **tmap** and **tidyverse** packages of R are currently installed in your R.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

## **8.3 Getting the Data Into R Environment**

In this section, you will learn how to bring a geospatial data and its associated attribute table into R environment. The geospatial data is in ESRI shapefile format and the attribute table is in csv fomat.

### **8.3.1 Import shapefile into r environment**

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "../Hands_on_Ex08/data/geospatial", 
                 layer = "Hunan")
```

### **8.3.2 Import csv file into r environment**

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package. The output is R dataframe class.

```{r}
hunan2012 <- read_csv("../Hands_on_Ex08/data/aspatial/Hunan_2012.csv")
```

### **8.3.3 Performing relational join**

The code chunk below will be used to update the attribute table of *hunan*’s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.
