---
title: "In-Class_exercise 4"
author: "chandru"
---

```{r}
pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)

```


##  Getting the Data Into R Environment**

In this section, you will learn how to bring a geospatial data and its associated attribute table into R environment. The geospatial data is in ESRI shapefile format and the attribute table is in csv fomat.

### **8.3. Import shapefile into r environment**

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan_sf <- st_read(dsn = "../In-class_Ex04/data/geospatial", 
                 layer = "Hunan")
```

### **8.3.2 Import csv file into r environment**

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package. The output is R dataframe class.

```{r}
hunan2012 <- read_csv("../In-class_Ex04/data/aspatial/Hunan_2012.csv")
```

### **8.3.3 Performing relational join**

The code chunk below will be used to update the attribute table of *hunan*’s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.

```{r}
hunan_sf <- left_join(hunan_sf,hunan2012) %>%
  select(1:3, 7,15,16,31,32)
```




```{r}

hunan_sp <- sf::as_Spatial(hunan_sf)

```


cross validation

```{r}

bw_cv<- bw.gwr(GDPPC ~ 1,
               data=hunan_sf,
               approach="CV",
               adaptive= FALSE,
               kernel = "bisquare",
               longlat= T)
bw_cv
```


```{r}

bw_AIC<- bw.gwr(GDPPC ~ 1,
               data=hunan_sf,
               approach="AIC",
               adaptive= FALSE,
               kernel = "bisquare",
               longlat= T)
bw_AIC
```

```{r}

gwstat<- gwss(
               data=hunan_sp,
               vars="GDPPC",
               bw = bw_AIC,
               adaptive= FALSE,
               kernel = "bisquare",
               longlat= T)

```

```{r}
gwstat_df<-as.data.frame(gwstat$SDF)
```

```{r}
hunan_gstat <- cbind(hunan_sf,gwstat_df)
```

