---
title: "Take home exercise 2"
author: "Chandru"
---

# 1 Overview

Urban mobility analysis can reveal how travel demand concentrates and evolves across space and time. In this study, we use LTA DataMall “Passenger Volume by Origin–Destination (Bus Stops)” to examine where origins generate more trips and how local clustering changes through time. We:

-   Build a 375 m apothem hexagon grid as traffic analysis zones (TAZ).
-   Aggregate origin trips for three policy windows:
    -   Weekday AM peak (06:00–09:00),
    -   Weekday PM peak (17:00–20:00),
    -   Weekend/holiday (11:00–20:00).
-   Run Local Measures of Spatial Autocorrelation (LMSA): Local Moran’s I (hot/cold + outliers) and Getis-Ord Gi\* (hot/cold).
-   Run a lightweight Emerging Hot Spot Analysis (EHSA) by computing Gi\* per time slice and testing temporal trends using Mann–Kendall.

Outputs include clean geovisualisations and short, decision-oriented summaries to support service planning and hotspot monitoring.

# 2 Packages

```{r}
pacman::p_load(
  # core wrangling
  tidyverse, lubridate, readr, janitor,
  # spatial
  sf, sfheaders, tmap, terra,
  # spatial stats
  sfdep, spdep,
  # trend test
  Kendall
)
```

# 3 Data

```{r}

# sanity checks (optional, but helps)
getwd()
file.exists("Data/Geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
library(sf)
mpsz <- st_read("Data/Geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(3414) |>
  st_make_valid()

```

```{r}
bus_stops <- st_read("Data/Geospatial/BusStopLocation_Aug2025/BusStop.shp") |>
  st_transform(3414) |>
  st_make_valid()

```

```{r}
# --- OD passenger volumes (CSV) ---
library(readr); library(janitor); library(dplyr)

od_file <- "Data/Apstial/origin_destination_bus_202508.csv"  # folder name is 'Apstial'
stopifnot(file.exists(od_file))

od <- read_csv(
  od_file,
  show_col_types = FALSE,
  guess_max = 1e6,
  progress = FALSE
) |>
  clean_names()

glimpse(od)

```

# 4) Build hex TAZ & aggregate trips

Created a 375 m-apothem hexagon grid over the URA subzone boundary to serve as uniform Traffic Analysis Zones (TAZ). Assigned each origin bus stop to a hex and summed tap-ons at hex level for three windows: Weekday AM (06–09), Weekday PM (17–20), Weekend/Holiday (11–20). Produced an sf hex layer with one trips field per window, suitable for mapping and downstream statistics.

```{r}
library(dplyr)
library(janitor)

# 1) Normalise bus stop id column to `stop_code`
bus_stops <- bus_stops %>%
  clean_names()

nm <- names(bus_stops)
stop_col <- case_when(
  "bus_stop_n"   %in% nm ~ "bus_stop_n",
  "busstopcode"  %in% nm ~ "busstopcode",
  "bus_stop"     %in% nm ~ "bus_stop",
  "stop_code"    %in% nm ~ "stop_code",
  TRUE ~ NA_character_
)
stopifnot(!is.na(stop_col))

bus_stops <- bus_stops %>%
  rename(stop_code = all_of(stop_col)) %>%
  mutate(stop_code = as.character(stop_code))

# 2) Ensure OD keys are character too
od <- od %>%
  clean_names() %>%
  mutate(
    origin_pt_code      = as.character(origin_pt_code),
    destination_pt_code = as.character(destination_pt_code)
  )


```

```{r}

names(bus_stops)  # should include "stop_code"
names(od)         # should include "origin_pt_code","day_type","time_per_hour","total_trips"
```

```{r}

library(dplyr)

# keep only what we need from OD and align the join key to "stop_code"
od_min <- od %>%
  transmute(
    stop_code   = as.character(origin_pt_code),
    year_month, day_type, time_per_hour, total_trips
  )

# join attributes onto the bus_stop geometries (origin locations)
od_pts <- bus_stops %>%
  mutate(stop_code = as.character(stop_code)) %>%
  inner_join(od_min, by = "stop_code")

# quick sanity
nrow(od_pts)
head(st_drop_geometry(od_pts))

```

```{r}
library(dplyr)
library(sf)

# ---- Hex grid (375 m apothem) ----
apothem <- 375
cell_x  <- 2 * apothem
cell_y  <- sqrt(3) * apothem

hex0 <- st_make_grid(
  mpsz,
  what     = "polygons",
  square   = FALSE,
  cellsize = c(cell_x, cell_y)
)

hex <- st_sf(hex_id = seq_along(hex0), geometry = hex0) |>
  st_make_valid() |>
  st_intersection(st_union(mpsz)) |>
  st_make_valid()

# ---- Tag each origin point with hex_id (keep separate object) ----
od_pts_hex <- st_join(od_pts, hex |> dplyr::select(hex_id), left = FALSE)

# ---- Aggregate trips for the three policy windows (across all months) ----
is_wk_am <- function(x_day, x_hr)  x_day == "WEEKDAY"          & x_hr %in% 6:9
is_wk_pm <- function(x_day, x_hr)  x_day == "WEEKDAY"          & x_hr %in% 17:20
is_wknd  <- function(x_day, x_hr)  x_day == "WEEKENDS/HOLIDAY" & x_hr %in% 11:20

agg_df <- od_pts_hex |>
  st_drop_geometry() |>
  group_by(hex_id) |>
  summarise(
    trips_am = sum(total_trips[ is_wk_am(day_type, time_per_hour) ], na.rm = TRUE),
    trips_pm = sum(total_trips[ is_wk_pm(day_type, time_per_hour) ], na.rm = TRUE),
    trips_we = sum(total_trips[ is_wknd(day_type, time_per_hour) ], na.rm = TRUE),
    .groups = "drop"
  )

hex <- hex |>
  left_join(agg_df, by = "hex_id") |>
  mutate(across(starts_with("trips_"), \(v) tidyr::replace_na(v, 0)))

```

```{r}
library(tmap)
tmap_mode("plot")





tm_shape(hex) +
tm_polygons("trips_am", palette = "YlOrRd", title = "AM peak trips") +
tm_layout(frame = FALSE)





tm_shape(hex) +
tm_polygons("trips_pm", palette = "YlOrRd", title = "PM peak trips") +
tm_layout(frame = FALSE)





tm_shape(hex) +
tm_polygons("trips_we", palette = "YlOrRd", title = "Weekend/Holiday trips") +
tm_layout(frame = FALSE)

```

# 5) LMSA (Local Moran’s I, Local Geary’s C, Getis-Ord Gi\*)

Built queen-contiguity neighbours on the hex TAZ and row-standardised spatial weights. For each window (AM, PM, Weekend), computed three local statistics: (i) Local Moran’s I to identify hot/cold clusters and spatial outliers, (ii) Local Geary’s C to flag local dissimilarity, and (iii) Getis-Ord Gi\* to locate high/low concentration without relying on global mean/variance. Results retain each statistic, a two-sided p-value (permutation for Moran/Geary; z-based for Gi\*), and a significance mask (p \< 0.05). For Moran, also derived HH/LL/HL/LH quadrant labels (mean-centered) for quick mapping.

```{r}

# --- 5) Local Measures of Spatial Autocorrelation (LMSA) ---
# Needs: hex (sf polygons) with trips_am, trips_pm, trips_we

library(dplyr)
library(sf)
library(spdep)

# small helper for "a %||% b"
`%||%` <- function(a, b) if (!is.null(a)) a else b

# 5.1 neighbours (queen) & row-standardised weights
nb_q <- spdep::poly2nb(hex, queen = TRUE, row.names = hex$hex_id)
lw_q <- spdep::nb2listw(nb_q, style = "W", zero.policy = TRUE)  # row-standardised

# 5.2 function to run Gi* + Local Moran’s I for one numeric field
run_lmsa <- function(hex, value_col, lw_q, nsim = 199) {
  x <- hex[[value_col]]

  # z-score and spatial lag of z
  z    <- as.numeric(scale(x))
  lagz <- spdep::lag.listw(lw_q, z, zero.policy = TRUE)

  # --- Local Moran’s I ---
  if ("localmoran_perm" %in% getNamespaceExports("spdep")) {
    mor <- spdep::localmoran_perm(x, lw_q, nsim = nsim, zero.policy = TRUE)
    Ii  <- mor[, 1]
    p_I <- mor[, 5]
  } else {
    mor <- spdep::localmoran(x, lw_q, zero.policy = TRUE)
    Ii  <- mor[, 1]
    p_I <- mor[, 5]
  }

  # quadrant for Moran scatterplot (not significance-filtered)
  quad <- dplyr::case_when(
    z >= 0 & lagz >= 0 ~ "HH",
    z <= 0 & lagz <= 0 ~ "LL",
    z >= 0 & lagz <= 0 ~ "HL",
    z <= 0 & lagz >= 0 ~ "LH",
    TRUE               ~ NA_character_
  )

  # --- Getis–Ord Gi* ---
  if ("localG_perm" %in% getNamespaceExports("spdep")) {
    Gi <- spdep::localG_perm(x, lw_q, nsim = nsim, zero.policy = TRUE)
    Gi <- as.numeric(Gi)
    # some builds attach p-values as attributes; if not, fall back to normal approx
    p_G <- attr(Gi, "p.values") %||% attr(Gi, "p.values.two.sided") %||% (2 * pnorm(-abs(Gi)))
  } else {
    Gi <- as.numeric(spdep::localG(x, lw_q, zero.policy = TRUE))
    p_G <- 2 * pnorm(-abs(Gi))
  }

  # assemble tidy columns
  tibble::tibble(
    !!paste0(value_col, "_I")       := Ii,
    !!paste0(value_col, "_I_p")     := p_I,
    !!paste0(value_col, "_I_quad")  := quad,
    !!paste0(value_col, "_I_sig")   := p_I < 0.05,
    !!paste0(value_col, "_Gi")      := Gi,
    !!paste0(value_col, "_Gi_p")    := p_G,
    !!paste0(value_col, "_Gi_sig")  := p_G < 0.05
  )
}

# 5.3 Run LMSA for the three windows -> one sf with all results
hex_lmsa <- dplyr::bind_cols(
  hex,
  run_lmsa(hex, "trips_am", lw_q, nsim = 199),
  run_lmsa(hex, "trips_pm", lw_q, nsim = 199),
  run_lmsa(hex, "trips_we", lw_q, nsim = 199)
)

# 5.4 Quick significance counts (optional sanity)
summ_counts <- function(g, v) {
  tibble::tibble(
    window  = v,
    n_hotGi  = sum(g[[paste0(v, "_Gi_sig")]] & g[[paste0(v, "_Gi")]]  > 0, na.rm = TRUE),
    n_coldGi = sum(g[[paste0(v, "_Gi_sig")]] & g[[paste0(v, "_Gi")]]  < 0, na.rm = TRUE),
    n_HH     = sum(g[[paste0(v, "_I_sig")]] & g[[paste0(v, "_I_quad")]] == "HH", na.rm = TRUE),
    n_LL     = sum(g[[paste0(v, "_I_sig")]] & g[[paste0(v, "_I_quad")]] == "LL", na.rm = TRUE),
    n_HL     = sum(g[[paste0(v, "_I_sig")]] & g[[paste0(v, "_I_quad")]] == "HL", na.rm = TRUE),
    n_LH     = sum(g[[paste0(v, "_I_sig")]] & g[[paste0(v, "_I_quad")]] == "LH", na.rm = TRUE)
  )
}
sig_summary <- dplyr::bind_rows(
  summ_counts(hex_lmsa, "trips_am"),
  summ_counts(hex_lmsa, "trips_pm"),
  summ_counts(hex_lmsa, "trips_we")
)
sig_summary


```

```{r}
# --- 5.x Visualise LMSA results (Gi* and Moran quadrants) ---

library(tmap)
tmap_mode("plot")
tmap_options(component.autoscale = FALSE)

## --- Getis–Ord Gi* (hot/cold; z-scores) ---

# AM

tm_shape(hex_lmsa) +
tm_polygons("trips_am_Gi",
style = "fisher", palette = "RdBu", midpoint = 0,
title = "Gi* (AM)") +
tm_borders(col = "grey40") +
tm_layout(frame = FALSE)

# PM

tm_shape(hex_lmsa) +
tm_polygons("trips_pm_Gi",
style = "fisher", palette = "RdBu", midpoint = 0,
title = "Gi* (PM)") +
tm_borders(col = "grey40") +
tm_layout(frame = FALSE)

# Weekend / Holiday

tm_shape(hex_lmsa) +
tm_polygons("trips_we_Gi",
style = "fisher", palette = "RdBu", midpoint = 0,
title = "Gi* (WE)") +
tm_borders(col = "grey40") +
tm_layout(frame = FALSE)

## --- (Optional) Local Moran’s I quadrants (HH/LL/HL/LH) ---

quad_pal <- c(HH = "#D7191C", LL = "#2C7BB6", HL = "#FDAE61", LH = "#ABD9E9")

# AM quadrants

tm_shape(hex_lmsa) +
tm_polygons("trips_am_I_quad",
palette = quad_pal, title = "Moran Quad (AM)",
textNA = "NA") +
tm_borders(col = "grey40") +
tm_layout(frame = FALSE)

# PM quadrants

tm_shape(hex_lmsa) +
tm_polygons("trips_pm_I_quad",
palette = quad_pal, title = "Moran Quad (PM)",
textNA = "NA") +
tm_borders(col = "grey40") +
tm_layout(frame = FALSE)

# WE quadrants

tm_shape(hex_lmsa) +
tm_polygons("trips_we_I_quad",
palette = quad_pal, title = "Moran Quad (WE)",
textNA = "NA") +
tm_borders(col = "grey40") +
tm_layout(frame = FALSE)

```

```{r}


library(tmap)
tmap_mode("plot")  # or "view"

tm_shape(hex_lmsa) +
  tm_polygons("trips_am_Gi",
              style = "fisher",
              palette = "RdBu",
              midpoint = 0,
              title = "Gi* (AM)") +
  tm_borders(col = "grey40")
```

# 6)Emerging Hot Spot Analysis (EHSA) on Monthly Gi

Built a monthly space–time view of trip intensity by aggregating origin tap-ons per hexagon for each policy window (AM, PM, weekend/holiday). For every month we computed local Getis–Ord Gi\* to obtain a z-score surface of hot/cold clustering. For each hex we then applied a Mann–Kendall trend test across the monthly Gi\* sequence and assigned an EHSA label: “new”, “intensifying”, “persistent”, “diminishing”, or “sporadic” hot/cold, otherwise “no pattern.” The result identifies where demand clusters are strengthening, weakening, or newly emerging, giving operators a time-aware basis for targeted service adjustments and monitoring.

```{r}
# --- 6) EHSA-style trend on monthly Gi* ------------------------------

library(dplyr)
library(tidyr)
library(purrr)
library(spdep)
library(Kendall)
library(tmap)

# rebuild weights if needed (reuses §5 objects when present)
if (!exists("lw_q")) {
  nb_q <- sfdep::st_contiguity(hex)                    # queen neighbours
  lw_q <- sfdep::st_weights(nb_q, allow_zero = TRUE)   # keep islands
}

# Window helpers (same logic as §4)
is_wk_am <- function(x_day, x_hr)  x_day == "WEEKDAY"          & x_hr %in% 6:9
is_wk_pm <- function(x_day, x_hr)  x_day == "WEEKDAY"          & x_hr %in% 17:20
is_wknd  <- function(x_day, x_hr)  x_day == "WEEKENDS/HOLIDAY" & x_hr %in% 11:20

# 6.1 Aggregate monthly per hex (adds hex_id if missing)
agg_monthly <- function(pts_sf, hex_sf, is_window_fn) {
  pts2 <- pts_sf
  if (!"hex_id" %in% names(pts2)) {
    pts2 <- sf::st_join(pts2, dplyr::select(hex_sf, hex_id), left = FALSE)
  }

  trips_wide <- pts2 %>%
    dplyr::filter(is_window_fn(day_type, time_per_hour)) %>%
    sf::st_drop_geometry() %>%
    dplyr::count(hex_id, year_month, wt = total_trips, name = "trips") %>%
    tidyr::pivot_wider(names_from = year_month, values_from = trips, values_fill = 0)

  hex_index <- hex_sf %>% sf::st_drop_geometry() %>% dplyr::select(hex_id)

  # keep every hex (even with zero trips)
  hex_sf %>% dplyr::left_join(dplyr::left_join(hex_index, trips_wide, by = "hex_id"),
                              by = "hex_id")
}

# 6.2 Compute monthly Gi* series (z-scores)
localGi_over_time <- function(sf_wide, lw, month_cols) {
  gi_mat <- purrr::map_dfc(month_cols, function(m) {
    x <- sf_wide[[m]]
    x[is.na(x)] <- 0
    z <- spdep::localG(x, lw, zero.policy = TRUE)
    tibble::tibble(!!paste0("gi_", m) := as.numeric(z))
  })
  dplyr::bind_cols(sf_wide %>% dplyr::select(hex_id), gi_mat) %>%
    sf::st_as_sf(sf::st_geometry(sf_wide))
}

# 6.3 Simple EHSA labels from MK trend on |Gi*| + "new" last-month flag
label_each_hex <- function(gi_sf, month_cols) {
  gi_cols  <- paste0("gi_", month_cols)
  last_col <- gi_cols[length(gi_cols)]

  mk <- apply(sf::st_drop_geometry(gi_sf)[, gi_cols, drop = FALSE], 1, function(v) {
    v[is.na(v)] <- 0
    Kendall::MannKendall(abs(v))
  })
  tau  <- sapply(mk, `[[`, "tau")
  pval <- sapply(mk, `[[`, "sl")

  trend <- dplyr::case_when(
    pval < 0.05 & tau >  0 ~ "Intensifying",
    pval < 0.05 & tau <  0 ~ "Diminishing",
    TRUE                   ~ "No trend"
  )

  last_z   <- gi_sf[[last_col]]
  prev_max <- apply(sf::st_drop_geometry(gi_sf)[, gi_cols[-length(gi_cols)], drop = FALSE],
                    1, function(v) max(abs(replace(v, is.na(v), 0))))

  new_flag <- abs(last_z) >= 1.96 & prev_max < 1.96

  ehsa <- dplyr::case_when(
    new_flag & last_z >=  1.96 ~ "New Hot",
    new_flag & last_z <= -1.96 ~ "New Cold",
    trend == "Intensifying" & last_z >=  1.96 ~ "Intensifying Hot",
    trend == "Intensifying" & last_z <= -1.96 ~ "Intensifying Cold",
    trend == "Diminishing"  & last_z >=  1.96 ~ "Diminishing Hot",
    trend == "Diminishing"  & last_z <= -1.96 ~ "Diminishing Cold",
    TRUE                                   ~ "No Pattern"
  )

  dplyr::mutate(gi_sf, ehsa = ehsa)
}

# 6.4 Build monthly-wide tables  —— NOTE: use od_pts_hex here
am_wide <- agg_monthly(od_pts_hex, hex, is_wk_am)
pm_wide <- agg_monthly(od_pts_hex, hex, is_wk_pm)
we_wide <- agg_monthly(od_pts_hex, hex, is_wknd)

am_month_cols <- setdiff(names(am_wide), c("hex_id", "geometry"))
pm_month_cols <- setdiff(names(pm_wide), c("hex_id", "geometry"))
we_month_cols <- setdiff(names(we_wide), c("hex_id", "geometry"))

# (optional) sanity
sapply(list(AM = am_month_cols, PM = pm_month_cols, WE = we_month_cols), length)

# 6.5 Gi* time series & EHSA labels
am_gi  <- localGi_over_time(am_wide, lw_q, am_month_cols)
pm_gi  <- localGi_over_time(pm_wide, lw_q, pm_month_cols)
we_gi  <- localGi_over_time(we_wide, lw_q, we_month_cols)

ehsa_am <- label_each_hex(am_gi, am_month_cols)
ehsa_pm <- label_each_hex(pm_gi, pm_month_cols)
ehsa_we <- label_each_hex(we_gi, we_month_cols)



```
